---
title: "README"
author: "aarvil"
date: "January 25, 2016"
output: html_document
---

## Getting and Cleaning Data Course Project


### Goals of run_analysis.R script:  
1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement. 
3. Uses descriptive activity names to name the activities in the data set
4. Appropriately labels the data set with descriptive variable names. 
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

### Datasets
| Dataset               | Description                       |
|-----------------------|-----------------------------------|
| TIDYDATASET.TXT       | New dataset consolidating training and test datasets with descriptive column names and activity labels. |
| TIDYDATASET-MEAN.TXT  | New dataset containing the mean of each measurements grouped by activity and subject                       |


### Original Data
Data usedon this
* https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip 

A full description is available at the site where the data was obtained:
* http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones 

### Additional information regarding variables
The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

* mean(): Mean value
* std(): Standard deviation

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

* gravityMean
* tBodyAccMean
* tBodyAccJerkMean
* tBodyGyroMean
* tBodyGyroJerkMean 

# Transformation
```{r}
library(dplyr)
library(stringr)
```
## STEP 1: LOAD DATA. -----------------------------------
```{r}
message("Loading data...")
```
### Read training dataset. 
```{r}
traindata <- read.table(file = "./UCI HAR Dataset/train/X_train.txt")
trainlabels <- read.table(file = "./UCI HAR Dataset/train/y_train.txt")
trainsubject <- read.table(file = "./UCI HAR Dataset/train/subject_train.txt")
```
### Read test dataset. 
```{r}
testdata <-read.table(file = "./UCI HAR Dataset/test/X_test.txt")
testlabels <- read.table(file = "./UCI HAR Dataset/test/y_test.txt")
testsubject <- read.table(file = "./UCI HAR Dataset/test/subject_test.txt")
```
### Read feature (column) name data. 
```{r}
features <- read.table(file = "./UCI HAR Dataset/features.txt", sep = " ")
```
### Read activity labels
```{r}
activitylabels <- read.table(file = "./UCI HAR Dataset/activity_labels.txt")
```
### Use data tables to make it easier to work with datasets. 
```{r}
traindata <- tbl_dt(traindata)
trainlabels <- tbl_dt(trainlabels)
trainsubject <- tbl_dt(trainsubject)
testdata <- tbl_dt(testdata)
testlabels <- tbl_dt(testlabels)
testsubject <- tbl_dt(testsubject)
features <- tbl_dt(features)
activitylabels <- tbl_dt(activitylabels)

message("Loading successful.")
```

## STEP 2: DATA TRANSFORMATION -------------------------------
```{r}
message("Starting data transformation...")
```
### 2.A: Apply descriptive column names to datasets using data from features.txt.

#### Coerce feature into characters in order to use them for column names. 
```{r}
features$V2 <- as.character(features$V2) 
```
#### Remove parentesis from variable names to increase readability. 
```{r}
features <- mutate(features, V3 = str_replace(features$V2, "\\(\\)", "")) 
```
#### Apply feature as column names to training and test datasets. 
```{r}
colnames(testdata) <- features$V3
colnames(traindata) <- features$V3
```

### 2.B: Extract measurements on the mean and standard deviation for each measurement.
```{r}
test2 <- select(testdata, contains("mean"), contains("std"))
train2 <- select(traindata, contains("mean"), contains("std"))
```
### 2.C: Add activity and subject variables into datasets
```{r}
test2 <- mutate(test2, activity = testlabels, subject = testsubject)
train2 <- mutate(train2, activity = trainlabels, subject = trainsubject) 
```
### 2.D: Merge the training and the test sets to create one data set.
```{r}
completedata <- rbind_list(test2, train2)
```
### 2.E: Replace activity id's with descriptive activity names.
```{r}
activity3 <- character(0)
activitylabels$V2 <- as.character(activitylabels$V2)

for(i in 1:length(completedata$activity)) {
      activity3 <- c(activity3, activitylabels$V2[completedata$activity[i]])
}
completedata$activity <- activity3

message("Transformation complete.")
```
### 2.F: Write data into a text file tidydata.txt in the working directory. 
```{r}
message("Writing data to file tidydataset.txt in working directory.")
write.table(completedata, file = "tidydataset.txt", row.names = FALSE)
message("tidydataset.txt file successfully created.")

```
## STEP 3: From final dataset, create a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r}
message("Creating new dataset with average of each variable for each activity and each subject.")
```
### Group data by activity and subject. 
```{r}
completedata <- group_by(completedata, activity, subject)
```
### Get the average of each variable for each activity and each subject. 
```{r}
meanbyactivityandsubject <- summarise_each(completedata, funs(mean))
```
### Write data into text file tidydataset-means.txt in working directory.
```{r}
message("Writing data to file tidydataset-means.txt in working directory.")
write.table(meanbyactivityandsubject, file = "tidydataset-means.txt", row.names = FALSE)      
message("tidydataset-means.txt file successfully created.")
```


## Environment Information
* platform       x86_64-w64-mingw32          
* arch           x86_64                      
* os             mingw32                     
* system         x86_64, mingw32             
* status                                     
* major          3                           
* minor          2.2                         
* year           2015                        
* month          08                          
* day            14                          
* svn rev        69053                       
* language       R                           
* version.string R version 3.2.2 (2015-08-14)
* nickname       Fire Safety    
